---
title: "tax-viburnum"
author: "JT Miller"
date: "2025-01-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(rgnparser)
library(tidyverse)
library(taxadb)
# Set up pathing to use gnparser
my_path <- Sys.getenv("PATH") # grab our path
Sys.setenv(PATH = paste0(my_path, "/home/millerjared/gnparser"))
```
Load in wcvp's backbone and the ncbi backbone
```{r}
# load in wcvp's taxonomic backbone
wcvp_backbone <- fread("/home/millerjared/blue_guralnick/millerjared/BoCP/data/raw/wcvp_names.csv")

# load in ncbi's taxonomic db
ncbi_names <- taxa_tbl("ncbi") # load in ncbi's names according to taxadb's last pull
ncbi_names <- ncbi_names %>%  collect() # collects into a df 
```
Remove non-plant data from the ncbi backbone 
Remove names that have questionable definitions, such as cf. 
```{r}
ncbi_names <- as.data.table(ncbi_names)
ncbi_plant_names <- ncbi_names[kingdom == "Viridiplantae"]
ncbi_plant_names <- ncbi_plant_names %>% 
  filter(!grepl(" cf", specificEpithet) | !grepl(" cf", scientificName) ) %>% 
  filter(!grepl(" cf  ", specificEpithet) | !grepl(" cf  ", scientificName))# note that specificEpithet is the aligned name. 
# ncbi_plant_names <- ncbi_plant_names %>% 
#   filter(!grepl("environmental sample", specificEpithet) | !grepl("environmental sample", scientificName))
ncbi_plant_names <- ncbi_plant_names %>% # we wont deal with hybrids for simplicity
  filter(!grepl(" x ", specificEpithet ) | !grepl(" x ", scientificName)) %>% 
  filter(!grepl(" sect ", specificEpithet) | !grepl(" sect ", scientificName)) %>% 
  filter(!grepl("/", specificEpithet) | !grepl("/", scientificName))

# remove further uncertainty: "/" or (Species A) etc. 
#test2 <- ncbi_plant_names %>% 
  #filter(!grepl("/", specificEpithet) | !grepl("/", scientificName)) %>% 
  #filter(grepl("species", specificEpithet) | grepl("species", scientificName)) %>%  
  
```

Standardize the two backbones for our purposes
```{r}
# select and rename fields of interest
ncbi_plant_names <- ncbi_plant_names[,.(genus, family, scientificName, taxonomicStatus, ncbiAcceptedNameUsageID = acceptedNameUsageID, ncbiAlignedName = specificEpithet)]

# parse the scientificName field into the name and authorship fields
ncbi_parsed_plant_names <- gn_parse_tidy(ncbi_plant_names$scientificName)  # break up names

# remove duplicates where present 
ncbi_parsed_plant_names <- ncbi_parsed_plant_names %>% 
  distinct(, .keep_all=TRUE)
# merge
ncbi_parsed_plant_names_df <- merge(ncbi_plant_names, ncbi_parsed_plant_names, by.x = 'scientificName', by.y = 'verbatim')
# we're not going to worry about complications related to names that CANNOT be parsed, so anything that gets truncated 
ncbi_parsed_plant_names_df <- ncbi_parsed_plant_names_df %>% 
  filter(!cardinality == 0) # 0 represents things that fail at parsing. 
# remove duplicates again due to merge
ncbi_parsed_plant_names_df <- ncbi_parsed_plant_names_df %>% 
  distinct(, .keep_all = TRUE)
# update field names for clarity
ncbi_parsed_plant_names_df <- ncbi_parsed_plant_names_df[,.(ncbiNameGenus = genus, ncbiNameFamily = family, ncbiVerbatimName = scientificName, ncbiName = canonicalfull, ncbiAuthor = authorship, ncbiTaxonomicStatus = taxonomicStatus, ncbiAcceptedNameUsageID, ncbiAlignedName)]



# ncbi is a bit difficult with parsing techniques as the names can be non-standard. One way about this is just taking those names out for now and attaching them as synonyms down the line to our wcvp names...REDACTED with cardinality rule
# ncbi_non_parsed_names <- ncbi_parsed_plant_names_df[is.na(ncbiName)]
# ncbi_parsed_plant_names_df <- ncbi_parsed_plant_names_df[!ncbiVerbatimName %in% ncbi_non_parsed_names$ncbiVerbatimName]

# note that in ncbi authority just means that there is an authorship attached to the name. we're going to standardize the names, then check how much multiple mapping matters in the subsequent chunk (to see if its just better to eliminate all of these names entirely.)

# create a relation that notes whether the authority is accepted or synonym & deal with other name details
ncbi_parsed_plant_names_dfx <- ncbi_parsed_plant_names_df %>% 
  filter(!ncbiTaxonomicStatus == "in-part") %>% # this maps to multiple aligned names by definition, which is a problem for alignment. 
  group_by(ncbiName) %>%  # ID multiple mapping names, as these will cause issues 
  mutate(ncbiNameMultipleMaps = ifelse(sum(ncbiTaxonomicStatus == "accepted") > 1, TRUE, FALSE)) %>% 
  ungroup() 
# check mult map impact in subsequent chunk, for now lets remove and move on. 

ncbi_parsed_plant_names_dfy <- ncbi_parsed_plant_names_dfx %>% 
  filter(ncbiNameMultipleMaps == FALSE) %>% # removed for simplicity. Check
  mutate(ncbiTaxonomicStatus = case_when(
    ncbiTaxonomicStatus == "synonym" ~ ncbiTaxonomicStatus, # return same relation
    ncbiTaxonomicStatus == "accepted" ~ ncbiTaxonomicStatus, # return same relation
    ncbiTaxonomicStatus == "common name" ~ "synonym", # alter field
    ncbiTaxonomicStatus == "equivalent name" ~ "synonym", # alter field
    ncbiTaxonomicStatus == "includes" ~ "synonym", # alter field
    ncbiTaxonomicStatus == "type material" ~ "synonym", # alter field 
    ncbiTaxonomicStatus == "genbank common name" ~ "synonym", # alter field
    ncbiTaxonomicStatus == "authority" & ncbiName == ncbiAlignedName & !is.na(ncbiAuthor) ~ "accepted with authorship", 
    ncbiTaxonomicStatus == "authority" & ncbiName != ncbiAlignedName & !is.na(ncbiAuthor) ~ "synonym with authorship"
  )) %>% 
  mutate(ncbiTaxonomicStatus = case_when( # double check as there are odd cases where these are missed
    ncbiName == ncbiAlignedName & !is.na(ncbiAuthor) ~ "accepted with authorship", 
    ncbiName != ncbiAlignedName & !is.na(ncbiAuthor) ~ "synonym with authorship",
    TRUE ~ ncbiTaxonomicStatus # else cases return 
  ))

# remove redundancy, if we have a synonym with authorship we dont need just the synonym, same logic for accepted names with authorship 
ncbi_parsed_plant_names_df2 <- ncbi_parsed_plant_names_dfy %>% 
  group_by(ncbiName) %>% 
  mutate(redundant = ifelse(
        # check for redundancy in synonym
          all(c("synonym", "synonym with authorship") %in% ncbiTaxonomicStatus) |
        # check for redundancy in accepted
          all(c("accepted", "accepted with authorship") %in% ncbiTaxonomicStatus),
        TRUE, # if there is redundancy
        FALSE # else return
    )) %>% 
  ungroup()

# when redundancy is TRUE, remove the name that has the least info (w/out author)
ncbi_parsed_plant_names_df3 <- ncbi_parsed_plant_names_df2 %>%
  filter(!(redundant == TRUE & is.na(ncbiAuthor))) # note that this isnt perfect, it'll remove names that fail parsing past the genus level, but im okay with this simplification. 
ncbi_parsed_plant_names_df3 <- as.data.table(ncbi_parsed_plant_names_df3)
# create a relational table
ncbi_relations <- ncbi_parsed_plant_names_df3 %>% 
  rename(ncbiAlignedGenus = ncbiNameGenus, ncbiAlignedFamily = ncbiNameFamily) %>% # fixing a slight correction and renaming for clarity
  group_by(ncbiAlignedName) %>% 
  mutate(ncbiAlignedAuthor = ifelse(
    ncbiTaxonomicStatus == "accepted with authorship", 
    ncbiAuthor,
    NA_character_ # deal with NAs
  )) %>% 
  fill(ncbiAlignedAuthor, .direction = "downup") %>% # fill ncbiAligned author for all rows sharing the same "ncbiAlignedName 
  mutate(ncbiAlignedNameStatus = ifelse(
    ncbiTaxonomicStatus == "accepted" | ncbiTaxonomicStatus == "accepted with authorship", 
    "accepted",
    NA_character_
  )) %>% 
  fill(ncbiAlignedNameStatus, .direction = "downup") %>% 
  ungroup() %>% 
  as.data.table()

ncbi_relations <- ncbi_relations[, .( # simplify fields to the ones that we need
  ncbiAlignedFamily,
  ncbiAlignedGenus,
  ncbiAcceptedNameUsageID,
  ncbiAlignedName,
  ncbiAlignedAuthor,
  ncbiAlignedNameStatus,
  ncbiName,
  ncbiNameTaxonomicStatus = ncbiTaxonomicStatus,
  ncbiAuthor

)]

# wcvp standardize
wcvp_names <- wcvp_backbone[, .(genus, family, plant_name_id, taxon_status, taxon_name, taxon_authors, accepted_plant_name_id)]

# create aligned name field (wcvp only) note we use backbone due to Viburnum filtering leaving out HETEROTYPIC synonyms. 
wcvp_accepted_mapping <- wcvp_backbone[taxon_status == "Accepted"]
wcvp_synonym_mapping <- wcvp_backbone[taxon_status %in% c("Synonym", "Illegitimate", "Invalid", "Misapplied", "Orthographic", "Unplaced")]
wcvp_relations <- merge(wcvp_accepted_mapping, wcvp_synonym_mapping, # Create a relational table for wcvp relations
                        by.x = "accepted_plant_name_id",
                        by.y = "accepted_plant_name_id", 
                        all.x = TRUE) #all.y = TRUE)
# Order and select fields for simplicity
wcvp_relations <- wcvp_relations[, .(
  wcvpAlignedGenus = genus.x,
  wcvpAlignedFamily = family.x,
  wcvpAcceptedNameUsageID = accepted_plant_name_id,
  wcvpAlignedName = taxon_name.x,
  wcvpAlignedNameAuthors = taxon_authors.x,
  wcvpAlignedNameStatus = taxon_status.x,
  wcvpName = taxon_name.y,
  wcvpNameAuthors = taxon_authors.y,
  wcvpNameStatus = taxon_status.y
)]
aligned_as_name <- wcvp_relations[, .(
  wcvpAlignedFamily,
  wcvpAlignedGenus,
  wcvpAcceptedNameUsageID,
  wcvpAlignedName,
  wcvpAlignedNameAuthors,
  wcvpAlignedNameStatus,
  wcvpName = wcvpAlignedName,
  wcvpNameAuthors = wcvpAlignedNameAuthors,
  wcvpNameStatus = wcvpAlignedNameStatus
)]

# Combine the original relations with the aligned names as names
wcvp_relations <- rbindlist(list(wcvp_relations, aligned_as_name), use.names = TRUE, fill = TRUE)

# Remove any duplicate rows to avoid redundancy
wcvp_relations <- unique(wcvp_relations)

# Removal of unplaced names. Note that those names that register as NA for wcvpAlignedName CANNOT arrive at a name due to being unplaced in wcvp's taxonomy. 
wcvp_relations <- wcvp_relations[wcvpNameStatus != "Unplaced" & !is.na(wcvpAlignedName)] # note that OR statement is because for whatever reason a name can be a synonym or otherwise attached to a unplaced name, however that name is not to be used regardless. 

```

**Remove for end product** Simplify to cases of just *Viburnum* for an example product. 
```{r}
# wcvp_relations2 <- wcvp_relations[wcvpAlignedGenus == "Viburnum" |wcvpAlignedFamily == "Adoxaceae" | grepl("Viburnum", wcvpName, ignore.case = TRUE)] # takes care of any situation where Viburnum shows up. 

wcvp_relations <- wcvp_relations %>% 
  filter(wcvpAlignedFamily == "Adoxaceae" | wcvpAlignedFamily == "Viburnaceae" | wcvpAlignedGenus == "Viburnum" | grepl("Viburnum", wcvpName, ignore.case = TRUE))

ncbi_relations <- ncbi_relations %>% 
  filter(ncbiAlignedFamily == "Adoxaceae" | ncbiAlignedFamily == "Viburnaceae" | ncbiAlignedGenus == "Viburnum" | grepl("Viburnum", ncbiName, ignore.case = TRUE))

# ncbi_relations <- ncbi_relations[ncbiAlignedGenus == "Viburnum" | ncbiAlignedFamily == "Adoxaceae" | grepl("Viburnum", ncbiName, ignore.case = TRUE)]
```

Identify cases where multiple mapping matters. 
```{r}
# cases where a wcvpName can go to different wcpvAlignedNames 
wcvp_relations <- wcvp_relations %>% 
  group_by(wcvpName) %>% 
  mutate(nNamePaths = n_distinct(wcvpAlignedName)) %>% 
  mutate(multipleNameAlignmentsPossible = ifelse(nNamePaths > 1, TRUE, FALSE))

# cases for ncbi, note we need to use a var earlier in the previous chunk to find this as it would cause issues to retain for author simplification. 
uniq_ncbi_names_w_multiple_alignment_path_issues <- ncbi_parsed_plant_names_dfx %>% 
  filter(ncbiNameMultipleMaps == TRUE) %>% 
  distinct(ncbiName, .keep_all = TRUE) %>% 
  nrow()

uniq_ncbi_names_wout_multiple_alignment_path_issues <- ncbi_parsed_plant_names_dfx %>% 
  filter(ncbiNameMultipleMaps == FALSE) %>% 
  distinct(ncbiName, .keep_all = TRUE) %>% 
  nrow()

uniq_names_w_multiple_alignment_path_issues <- wcvp_relations %>% 
 filter(multipleNameAlignmentsPossible == TRUE) %>% 
 distinct(wcvpName, .keep_all = TRUE) %>% 
 nrow()


uniq_names_wout_multiple_alignment_path_issues <- wcvp_relations %>% 
 filter(multipleNameAlignmentsPossible == FALSE) %>% 
 distinct(wcvpName, .keep_all = TRUE) %>% 
 nrow()

# calc the % of data effected by multiple name alignment issues. 
(uniq_names_w_multiple_alignment_path_issues/uniq_names_wout_multiple_alignment_path_issues) * 100

(uniq_ncbi_names_w_multiple_alignment_path_issues/uniq_ncbi_names_wout_multiple_alignment_path_issues) * 100

# The percentage is >5%, therefore I feel comfortable removing these given names.
wcvp_relations <- wcvp_relations %>% 
  filter(!(multipleNameAlignmentsPossible == TRUE & !wcvpNameStatus == "Accepted")) # remove, unless its an accepted name as that would cause issues. (We're ignoring this for simplicity.)

ncbi_relations <- ncbi_relations # these names have already been removed. 
  
```

Combine Alignments, identify where ncbi IDs will cause issues, and backfill ncbiIDs for all synonyms
```{r}
# First create a check to see if ncbi accepted names are ever in the synonym category of wcvp names, as we need to lump them together 
ncbi_relations_check <- ncbi_relations %>% 
  filter(!is.na(ncbiNameTaxonomicStatus)) %>% # not entirely sure how this snuck in there
  #rename(name = ncbiName, nameStatus = ncbiNameTaxonomicStatus) %>% 
  mutate(source = "ncbi") %>% 
  rename(ncbiNameStatus = ncbiNameTaxonomicStatus) %>% 
  mutate(ncbiNameStatus = case_when(
    ncbiNameStatus == "accepted with authorship" ~ "Accepted",
    ncbiNameStatus == "synonym with authorship" ~ "Synonym", 
    ncbiNameStatus == "accepted" ~ "Accepted", 
    ncbiNameStatus == "synonym" ~ "Synonym"
  )) %>% 
  mutate(ncbiAlignedNameStatus = str_to_title(ncbiAlignedNameStatus))

wcvp_relations_check <- wcvp_relations %>% 
  mutate(source = "wcvp")

# deal with a small example
wcvp_relations_test <- wcvp_relations_check %>%  
  filter(wcvpAlignedName == "Viburnum erubescens") %>% 
  select(wcvpAlignedName, wcvpAlignedNameStatus, wcvpName, wcvpNameStatus, source)

ncbi_relations_test <- ncbi_relations_check %>% 
  filter(ncbiName == "Viburnum burmanicum" | ncbiAlignedName == "Viburnum burmanicum") %>% 
  select(ncbiAlignedName, ncbiAlignedNameStatus, ncbiName, ncbiNameStatus, source, ncbiAcceptedNameUsageID) 

test_wcvp <- data.frame(wcvpAlignedName = c("Adoxa corydalifolia", "Adoxa corydalifolia", "Adoxa corydalifolia", "Adoxa corydalifolia var. corydalifolia", "Adoxa corydalifolia var. madeupicus", "Adoxa corydalifolia var. madeupicus"),
                        wcvpAlignedNameStatus = c("Accepted", "Accepted", "Accepted", "Accepted", "Accepted", "Accepted"),
                        wcvpName = c("Sinadoxa corydalifolia", "Adoxa corydalifolia", "Bogus lexicus", "Adoxa corydalifolia var. corydalifolia", "Adoxa corydalifolia var. madeupicus", "Adoxaquilt made var. up"),
                        wcvpNameStatus = c("Synonym", "Accepted", "Synonym", "Accepted", "Accepted", "Accepted"), 
                        source = c("wcvp", "wcvp", "wcvp", "wcvp", "wcvp", "wcvp"))

test_ncbi <- data.frame(
  ncbiAlignedName = c("Adoxa corydalifolia", "Sinadoxa corydalifolia", "Sinadoxa corydalifolia var. madeitup"),
  ncbiAlignedNameStatus = c("Accepted", "Accepted", "Accepted"),
  ncbiName = c("Adoxa corydalifolia", "Sinadoxa corydalifolia", "Sinadoxa corydalifolia var. madeitup"),
  ncbiNameStatus = c("Accepted", "Accepted", "Accepted"), 
  source = c("ncbi","ncbi", "ncbi"),
  ncbiAcceptedNameUsageID = c("NCBI:111", "NCBI:222", "NCBI:333")
)

# deal with infraspecifics by making a parentName
test_wcvp <- test_wcvp %>% 
  mutate(wcvpParentAlignedName = word(wcvpAlignedName, 1,2)) %>% 
  select(wcvpAlignedName, wcvpParentAlignedName, wcvpAlignedNameStatus, wcvpName, wcvpNameStatus, source)
test_ncbi <- test_ncbi %>% 
   mutate(ncbiParentAlignedName = word(ncbiAlignedName, 1,2)) %>% 
   select(ncbiAlignedName, ncbiParentAlignedName, ncbiAlignedNameStatus, ncbiName, ncbiNameStatus, source, ncbiAcceptedNameUsageID)
# Perform a full join on the names
merged_df <- full_join(test_wcvp, test_ncbi, by = c("wcvpName" = "ncbiName"), suffix = c("", "_ncbi"), keep = TRUE)


Excellent. Second change I need to do: I would like to manipulate the ncbiPossibleIDs field a bit. If there are multiple values (not NA), then I would like to prioritize which id comes first. It should be the case that if the accepted name matches wcvp's accepted name, then that ID goes first, anything else after.

To achieve the prioritization of ncbiPossibleIDs based on whether the accepted name matches wcvp's accepted name, we can follow these steps:

    Preprocess the Dataframes to Create Parent Names.
    Perform the Full Join.
    Process the Merged Dataframe:
        Determine the aligned parent name and aligned name.
        Identify source conflicts.
        Add a source field.
        Prioritize ncbiPossibleIDs based on the match with wcvp's accepted name.
```
Here's the updated code to achieve this:
```{r}
# Load necessary libraries
library(dplyr)
library(stringr)

# Create the dataframes
test_wcvp <- data.frame(
  wcvpAlignedName = c("Adoxa corydalifolia", "Adoxa corydalifolia", "Adoxa corydalifolia", "Adoxa corydalifolia var. corydalifolia", "Adoxa corydalifolia var. madeupicus", "Adoxa corydalifolia var. madeupicus", "Whatevericus mer"),
  wcvpAlignedNameStatus = c("Accepted", "Accepted", "Accepted", "Accepted", "Accepted", "Accepted", "Accepted"),
  wcvpName = c("Sinadoxa corydalifolia", "Adoxa corydalifolia", "Bogus lexicus", "Adoxa corydalifolia var. corydalifolia", "Adoxa corydalifolia var. madeupicus", "Adoxaquilt made var. up", "Whatevericus mer"),
  wcvpNameStatus = c("Synonym", "Accepted", "Synonym", "Accepted", "Accepted", "Accepted", "Accepted"), 
  source = c("wcvp", "wcvp", "wcvp", "wcvp", "wcvp", "wcvp", "wcvp")
)

test_ncbi <- data.frame(
  ncbiAlignedName = c("Adoxa corydalifolia", "Sinadoxa corydalifolia", "Sinadoxa corydalifolia var. madeitup"),
  ncbiAlignedNameStatus = c("Accepted", "Accepted", "Accepted"),
  ncbiName = c("Adoxa corydalifolia", "Sinadoxa corydalifolia", "Sinadoxa corydalifolia var. madeitup"),
  ncbiNameStatus = c("Accepted", "Accepted", "Accepted"), 
  source = c("ncbi","ncbi", "ncbi"),
  ncbiAcceptedNameUsageID = c("NCBI:111", "NCBI:222", "NCBI:333")
)

# Preprocess the dataframes to create parent names
test_wcvp <- test_wcvp %>%
  mutate(wcvpParentAlignedName = str_extract(wcvpAlignedName, "^\\w+ \\w+")) %>%
  select(wcvpAlignedName, wcvpParentAlignedName, wcvpAlignedNameStatus, wcvpName, wcvpNameStatus, source)

test_ncbi <- test_ncbi %>%
  mutate(ncbiParentAlignedName = str_extract(ncbiAlignedName, "^\\w+ \\w+")) %>%
  select(ncbiAlignedName, ncbiParentAlignedName, ncbiAlignedNameStatus, ncbiName, ncbiNameStatus, source, ncbiAcceptedNameUsageID)

# Perform a full join on the names
merged_df <- full_join(test_wcvp, test_ncbi, by = c("wcvpName" = "ncbiName"), suffix = c("", "_ncbi"), keep = TRUE)

# Process the merged dataframe
output_df <- merged_df %>%
  mutate(
    alignedParentName = coalesce(wcvpParentAlignedName, ncbiParentAlignedName),
    alignedName = coalesce(wcvpAlignedName, ncbiAlignedName),
    name = coalesce(wcvpName, ncbiName),
    sourceConflict = ifelse(!is.na(wcvpName) & !is.na(ncbiName) & wcvpNameStatus == "Synonym" & ncbiNameStatus == "Accepted", TRUE, FALSE),
    source = ifelse(!is.na(wcvpName), "wcvp", "ncbi")
  ) %>%
  group_by(alignedParentName) %>%
  mutate(
    ncbiPossibleIDs = ifelse(all(is.na(ncbiAcceptedNameUsageID)), NA, {
      # Extract unique IDs, ensuring NA values are omitted
      unique_ids <- unique(na.omit(ncbiAcceptedNameUsageID))
      # Prioritize the ID that matches wcvp's accepted name
      if (any(!is.na(wcvpAlignedName) & !is.na(ncbiAlignedName) & ncbiAlignedName == wcvpAlignedName, na.rm = TRUE))
 {
        wcvp_id <- ncbiAcceptedNameUsageID[ncbiAlignedName == wcvpAlignedName]
        wcvp_id <- unique(wcvp_id[!is.na(wcvp_id)])
        other_ids <- unique_ids[!unique_ids %in% wcvp_id]
        paste(c(wcvp_id, other_ids), collapse = "|")
      } else {
        paste(unique_ids, collapse = "|")
      }
    }),
    mult_ncbi_ids = n_distinct(ncbiAcceptedNameUsageID, na.rm = TRUE) > 1,
    ncbiAcceptedNameUsageID = ifelse(is.na(ncbiAcceptedNameUsageID), first(ncbiAcceptedNameUsageID, na_rm = TRUE), ncbiAcceptedNameUsageID)
  ) %>%
  ungroup() %>%
  select(alignedName, alignedParentName, name, sourceConflict, wcvpName, ncbiName, ncbiAcceptedNameUsageID, mult_ncbi_ids, ncbiPossibleIDs, source)



```

Changed 1/29/2025. Fixed issues related to multiple ncbi ids. 
```{r}
# Preprocess the dataframes to create parent names
wcvp_relations_check <- wcvp_relations_check %>%
  mutate(wcvpParentAlignedName = str_extract(wcvpAlignedName, "^\\w+ \\w+")) %>%
  select(wcvpAlignedName, wcvpParentAlignedName, wcvpAlignedNameStatus, wcvpName, wcvpNameStatus, source) %>% 
  filter(str_count(wcvpAlignedName, "\\w+") > 1) # removes genera level alignment

ncbi_relations_check <- ncbi_relations_check %>%
  mutate(ncbiParentAlignedName = str_extract(ncbiAlignedName, "^\\w+ \\w+")) %>%
  select(ncbiAlignedName, ncbiParentAlignedName, ncbiAlignedNameStatus, ncbiName, ncbiNameStatus, source, ncbiAcceptedNameUsageID) %>% 
  filter(str_count(ncbiAlignedName, "\\w+") > 1)

### Executing updated name alignmnet
# Perform a full join on the names
merged_df <- full_join(wcvp_relations_check, ncbi_relations_check, by = c("wcvpName" = "ncbiName"), suffix = c("", "_ncbi"), keep = TRUE)

# Process the merged dataframe
output_df <- merged_df %>%
  mutate(
    alignedParentName = coalesce(wcvpParentAlignedName, ncbiParentAlignedName),
    alignedName = coalesce(wcvpAlignedName, ncbiAlignedName),
    name = coalesce(wcvpName, ncbiName),
    sourceConflict = ifelse(!is.na(wcvpName) & !is.na(ncbiName) & wcvpNameStatus == "Synonym" & ncbiNameStatus == "Accepted", TRUE, FALSE),
    source = ifelse(!is.na(wcvpName), "wcvp", "ncbi")
  ) %>%
  group_by(alignedParentName) %>%
  mutate(
    ncbiPossibleIDs = ifelse(all(is.na(ncbiAcceptedNameUsageID)), NA, {
      # Extract unique IDs, ensuring NA values are omitted
      unique_ids <- unique(na.omit(ncbiAcceptedNameUsageID))
      # Prioritize the ID that matches wcvp's accepted name
     if (any(!is.na(wcvpAlignedName) & !is.na(ncbiAlignedName) & ncbiAlignedName == wcvpAlignedName, na.rm = TRUE))
 {
        wcvp_id <- ncbiAcceptedNameUsageID[ncbiAlignedName == wcvpAlignedName]
        wcvp_id <- unique(wcvp_id[!is.na(wcvp_id)])
        other_ids <- unique_ids[!unique_ids %in% wcvp_id]
        paste(c(wcvp_id, other_ids), collapse = "|")
      } else {
        paste(unique_ids, collapse = "|")
      }
    }),
    mult_ncbi_ids = n_distinct(ncbiAcceptedNameUsageID, na.rm = TRUE) > 1,
    ncbiAcceptedNameUsageID = ifelse(is.na(ncbiAcceptedNameUsageID), first(ncbiAcceptedNameUsageID, na_rm = TRUE), ncbiAcceptedNameUsageID)
  ) %>%
  ungroup() %>%
  select(alignedName, alignedParentName, name, sourceConflict, wcvpName, ncbiName, ncbiAcceptedNameUsageID, mult_ncbi_ids, ncbiPossibleIDs, source)



```





# REPEX
```{r}

# Combine the two tables with backfilling of ncbiAcceptedNameUsageID for wcvp names
combined <- wcvp_relations %>%
  rename(name = wcvpName, nameStatus = wcvpNameStatus) %>%
  mutate(source = "wcvp", ncbiAcceptedNameUsageID = NA) %>%
  bind_rows(
    ncbi_relations %>%
      rename(name = ncbiName, nameStatus = ncbiNameTaxonomicStatus, wcvpAlignedName = ncbiAlignedName,
             wcvpAlignedNameStatus = ncbiAlignedNameStatus) %>%
      mutate(source = "ncbi")
  ) %>%
  # Backfill ncbiAcceptedNameUsageID for wcvp entries by matching wcvpAlignedName and ncbiAlignedName
  left_join(ncbi_relations %>% select(ncbiAlignedName, ncbiAcceptedNameUsageID), 
            by = c("wcvpAlignedName" = "ncbiAlignedName")) %>%
  mutate(ncbiAcceptedNameUsageID = coalesce(ncbiAcceptedNameUsageID.x, ncbiAcceptedNameUsageID.y)) %>% 
  distinct(.keep_all = TRUE) %>% 
  mutate(word_count = str_count(wcvpAlignedName, "\\s+") + 1)
combined <- combined %>% 
   bind_rows(combined %>% 
              filter(word_count > 2) %>% 
              distinct(wcvpAlignedName, .keep_all = TRUE)) %>% 
   mutate(wcvpAlignedNameParent = ifelse(word_count > 2, 
                                     stringr::word(wcvpAlignedName, 1, 2), 
                                     wcvpAlignedName)) %>% 
  distinct(.keep_all = TRUE)

# Clean up duplicate columns
combined <- combined %>%
  select(-ncbiAcceptedNameUsageID.x, -ncbiAcceptedNameUsageID.y)

# Identify multiple ncbiAcceptedNameUsageIDs
result <- combined %>%
  group_by(wcvpAlignedName) %>%
  mutate(multiple_ncbi_ids = n_distinct(ncbiAcceptedNameUsageID[!is.na(ncbiAcceptedNameUsageID)]) > 1) %>%
  ungroup() %>% 
  distinct(.keep_all = TRUE)

result <- result %>% 
  group_by(wcvpAlignedName, name) %>% 
  mutate(source = factor(source, levels = c("wcvp", "ncbi"))) %>% 
  arrange(source) %>% # prioritize wcvp over ncbi
  filter(row_number() == 1) %>% 
  ungroup() %>% 
  mutate(wcvpAlignedNameStatus = str_to_title(wcvpAlignedNameStatus)) %>% 
  mutate(nameStatus = str_to_title(nameStatus))

# final step, deal with infraspecific information. We're collasping to the species level, regardless. We'll include any ncbi info from these.
# final_relations <- result %>% 
#   mutate(word_count = str_count(wcvpAlignedName, "\\s+") + 1)
# 
# test <- final_relations %>% 
#   bind_rows(final_relations %>% 
#               filter(word_count > 2) %>% 
#               distinct(wcvpAlignedName, .keep_all = TRUE)) %>% 
#               #mutate(synonym = wcvpAlignedName, 
#                 #     synonym_taxon_authors = wcvpAlignedNameAuthors, 
#                  #    synonym_powo_id = wcvpAcceptedNameUsageID)) %>% 
#   mutate(wcvpAlignedNameParent = ifelse(word_count > 2, 
#                                      stringr::word(wcvpAlignedName, 1, 2), 
#                                      wcvpAlignedName)) %>% 
#   distinct(.keep_all = TRUE)
# 
# # check for multiple mapping after reduction to wcvpAlignedNameParent
# test <- test %>% 
#   group_by(wcvpAlignedNameParent) %>% 
#   mutate(multipleParentNameMaps = n_distinct())

```
Redone to deal with names that are synonyms 
```{r}
library(dplyr)
library(stringr)

# Combine the two tables with backfilling of ncbiAcceptedNameUsageID for wcvp names
combined <- wcvp_relations %>%
  rename(name = wcvpName, nameStatus = wcvpNameStatus) %>%
  mutate(source = "wcvp", ncbiAcceptedNameUsageID = NA) %>%
  bind_rows(
    ncbi_relations %>%
      rename(name = ncbiName, nameStatus = ncbiNameTaxonomicStatus, wcvpAlignedName = ncbiAlignedName,
             wcvpAlignedNameStatus = ncbiAlignedNameStatus) %>%
      mutate(source = "ncbi")
  ) %>%
  # Backfill ncbiAcceptedNameUsageID for wcvp entries by matching wcvpAlignedName and ncbiAlignedName
  left_join(ncbi_relations %>% select(ncbiAlignedName, ncbiAcceptedNameUsageID), 
            by = c("wcvpAlignedName" = "ncbiAlignedName")) %>%
  mutate(ncbiAcceptedNameUsageID = coalesce(ncbiAcceptedNameUsageID.x, ncbiAcceptedNameUsageID.y)) %>%
  distinct(.keep_all = TRUE)

# Add step to check if ncbiAlignedName is an optional synonym under wcvp
combined <- combined %>%
  left_join(
    wcvp_relations %>%
      filter(nameStatus == "synonym") %>%
      select(optionalSynonym = name, wcvpAlignedName), 
    by = c("ncbiAlignedName" = "optionalSynonym")
  ) %>%
  mutate(
    wcvpAlignedName = coalesce(wcvpAlignedName.x, wcvpAlignedName.y, wcvpAlignedName)
  ) %>%
  select(-wcvpAlignedName.x, -wcvpAlignedName.y)

# Continue with word count and parent alignment logic
combined <- combined %>%
  mutate(word_count = str_count(wcvpAlignedName, "\\s+") + 1) %>%
  bind_rows(
    combined %>%
      filter(word_count > 2) %>%
      distinct(wcvpAlignedName, .keep_all = TRUE)
  ) %>%
  mutate(wcvpAlignedNameParent = ifelse(
    word_count > 2,
    stringr::word(wcvpAlignedName, 1, 2),
    wcvpAlignedName
  )) %>%
  distinct(.keep_all = TRUE)

# Clean up duplicate columns
combined <- combined %>%
  select(-ncbiAcceptedNameUsageID.x, -ncbiAcceptedNameUsageID.y)

# Identify multiple ncbiAcceptedNameUsageIDs
result <- combined %>%
  group_by(wcvpAlignedName) %>%
  mutate(multiple_ncbi_ids = n_distinct(ncbiAcceptedNameUsageID[!is.na(ncbiAcceptedNameUsageID)]) > 1) %>%
  ungroup() %>%
  distinct(.keep_all = TRUE)

# Prioritize sources and clean statuses
result <- result %>%
  group_by(wcvpAlignedName, name) %>%
  mutate(source = factor(source, levels = c("wcvp", "ncbi"))) %>%
  arrange(source) %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  mutate(
    wcvpAlignedNameStatus = str_to_title(wcvpAlignedNameStatus),
    nameStatus = str_to_title(nameStatus)
  )

```


